---
title: 'Week 5: R Functions Lab'
author: "Chantal Rabay"
date: "2/7/2022"
output: pdf_document
---

Here we will write a function to grade some student homework.

Simple input example: vector of student scores
```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```
>Question 1. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. 

```{r}
student1
```
The regular average will be returned by the 'mean()' function.
```{r}
mean(student1)
```
Finding the position of the minimum value in a vector using 'which.min()'
```{r}
which.min(student1)
```

```{r}
student1[8]
```
Returning the minimum value.
```{r}
student1[ which.min(student1)]
```
To get everything but the min value.
```{r}
student1[-which.min(student1)]
```
Take the mean.
```{r}
#First solution
mean( student1[-which.min(student1)])
```
Changing NA values to zero. Using the 'is.na()' function.
```{r}
student2
is.na(student2)
```

```{r}
x <- student2
x

x[is.na(x)] <- 0
x
mean(x)
```
Find the average score for student 3
```{r}
x <- student3
x[is.na(x)] <- 0
mean( x[ -which.min(x)])
```
### Making the function
Using the working snippet from above I will make a function
```{r}
grade <- function(x) {
  x[is.na(x)] <- 0
  mean( x[ -which.min(x)])
}
```
Example using function:
```{r}
grade(student1)
grade(student2)
grade(student3)
```
>Your final function should be adquately explained with **code comments** and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
#' Calculate average scores for a vector of homework scores
#' Dropping the lowest single score. Missing values will be treated as zero scores
#' @param x Numeric vector of homework scores
#'
#' @return Average score
#' @export
#'
#' @examples
#'    student <- c(100, NA, 90, 80)
#'    grade(student)
grade <- function(x) {
  # This Maps NA missing homework values to zero 
  # Missing homeworks score zero 
  x[is.na(x)] <- 0
  # We exclude the lowest score homework 
  mean( x[ -which.min(x)])
}
```
###**Grading the whole class** using the gradebook:
```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names=1)
```

```{r}
apply(gradebook,1,grade)
```
> Question 2. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

```{r}
results <- apply(gradebook,1,grade)
which.max(results)
```

> Question 3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained
the lowest scores overall?  [2pts]

Here we want to calculate a summary stat for each column of the gradebook. Which stat should we use.
```{r}
#Let's try average 
hw.ave <- (apply(gradebook, 2, mean, na.rm=TRUE))
which.min(hw.ave)
```
Plot data to see and compare medians.
```{r}
boxplot(gradebook)
```
> Question 3 Answer
**Homework two is the most difficult question.**

> Question 4. Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
gradebook[is.na(gradebook)] <- 0
cor(results,gradebook$hw5)
```
```{r}
apply(gradebook,2,cor,x=results)
```
HW 2 was least predictive, HW 5 was most predictive of overall score.

> Question 5. Make sure you save your Rmarkdown document and can click the “Knit” button to generate a PDF foramt report without errors. Finally, submit your PDF to gradescope. [1pt]


```{r}

```

```{r}
ncol(gradebook)
nrow(gradebook)
dim(gradebook)
```
```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names=1)
```

```{r}
mean(student3, na.rm=TRUE)
mean(na.omit(student3))
mean(student3)
mean(is.na(student3))
```
```{r}
sum(is.na(student2))
```

